### 2015-03-29 markdown/manuscript.tex
first draft:

More than a decade ago, Rick Riolo, Bill Worzel and I were working on a consulting project together that involved genetic programming. As we chatted one day, Rick was asked what he'd most like to see as part of the research program of GP ``moving forward''. Rick's answer informs this contribution, as well as much of my professional work with GP in the years since.

As I recall, Rick said he'd like work to focus on the ``symptoms'' we often see in evolutionary search processes: premature convergence, failure to improve, catastrophic lack of diversity, mysterious things like that sense we all get when we look at results that suggests \emph{we made bad choices} of some sort.

@@

This contribution aims squarely at that sensibility which sees the ``symptoms'' and ``pathologies'' as something we as a research community should be trying to stamp out \emph{before} letting users run their own GP systems. Instead, I'll argue that because GP differs qualitatively and philosophically from most other machine learning approaches, and also from the broader lay understanding of the ``automated design'' and exploratory research, much of the value GP offers is lost or ignored when we approach it as something to be ``cleaned up''.

@@

I would much rather we faced the fact that this shocked response to \emph{seeing what's happening} is inevitable---not just in the form of unwitting surprise from silly lay ``customers'' in an application project, but also from ourselves when we undertake any \emph{interesting} GP project, whether the project is ``theoretical'' or ``practical'', ``small'' or ``large''.

Any \emph{interesting} project will resist our best expectations. Now and then we should question whether our habitual response to that inevitable resistance is in any sense effective.

@@

Braithwaite's exercise feels subjectively \emph{harder} for more advanced programmers honing their development form; he suggests that novice programmers haven't learned ingrained but questionable habits, and haven't identified ``shortcuts'' that ``simplify'' the practices. In the same way, this exercise will feel \emph{most} artificial and restrictive to those of us with the most experience with GP. But like the martial arts exercises from software \emph{kata} were inspired, it isn't intended to be simple or even pleasant for the participants.

I intend it to expose habits I'll tactlessly call ``psueudo-GP'' among those of us who have learned through the years to think it's cheap and painless to \emph{just shut it off and start over} when things start to get strange in the course of a GP project. As an important side-effect, it helps surface that philosophical problem I alluded to above. But I have found the most useful result is how strongly and immediately it suggests tools with which one can address Rick's decade-old wish. It forces the participant to formally identify ``pathology'' and ``symptom'' before allowing them to attempt a ``fix''\ldots{} and even that must done with an intentionally limited set of tools.

@@

\section{How we treat GP, and how it treats us in turn}\hypertarget{how-we-treat-gp-and-how-it-treats-us-in-turn}{}\label{how-we-treat-gp-and-how-it-treats-us-in-turn}

Genetic Programming\footnote{And not just Genetic Programming as such, but also the broader discipline to which I claim it belongs and which is not obliged to be either ``genetic'' or ``programming''. I prefer to call this looser collection of practices ``generative processing'', and will also abbreviate it ``GP''; assume I mean the latter in every case.} embodies a very particular \emph{stance} towards the scientific and engineering work of modeling, design, analysis and optimization. I increasingly suspect the resistance we've all recounted towards GP from our prospective technical and lay audience has little to do with our technical results, but rather arises from uncertainty among that audience with GP's very particular ``way of working''.

There is a tacit assumption, even among GP theorists and practitioners, that science and engineering are only ``rigorous'' when they proceed through a sequence of ordered phases from planning to implementation. The ``scientific method'' is most often represented something along the lines of

\begin{quote}
\begin{enumerate}
\item conceptualization; 2. planning; 3. design; 4. architecture; 5. implementation; 6. testing; 7. debugging
\end{enumerate}
\end{quote}

I am sure very few scientists or engineers of my acquaintance would admit any \emph{real} project (in history) has ever followed this narrative arc in a literal sense. But that story nonetheless informs and constrains much of our work lives, from fund-raising to publishing reports:

\begin{quote}
Because of the body of published work, an insight was had. The insight was framed as a formal hypothesis. The hypothesis (and current Best Statistical Practices) suggested an experimental design, one familiar and obvious to any in Our Discipline. The experimental design was undertaken, the data were collected, the hypothesis duly tested, and now we can be confident of its veracity because\ldots{} well, you just heard me say ``Best Statistical Practices'', right?
\end{quote}

Under trivial term substitutions---``cost--benefit analysis'' and ``requirements document'' for ``hypotheses'' and ``experimental design'', for example---the same narrative can be used to describe almost any institutional project management and public policy planning process as well. The flow in every case is essentially from \emph{vision} to \emph{plan}, \emph{plan} to \emph{implementation}, \emph{implementation} to \emph{verification}, and \emph{verification} to \emph{validation}.

Of course, nobody ``really believes'' this narrative who has ever done any of the work. It is a matter for another day to draw parallels with the social construction of religious belief.\footnote{Paul Veyne's excellent \emph{Did the Greeks Believe in Their Myths?} might be an interesting starting point, though.} There have been numerous philosophical challenges to this artificial narrative of course, from Peirce and Dewey nearly a century ago, to Kuhn and Lakatos in the 1970s, and many more to be found in the Table of Contents in any Philosophy of Science text.

One in particular is my focus today: Andrew Pickering.

\subsection{Pickering's ``Mangle of Practice''}\hypertarget{pickerings-mangle-of-practice}{}\label{pickerings-mangle-of-practice}

Andrew Pickering's \emph{Mangle of Practice} is a decade old, but surprisingly little-known outside the field of Science Studies. His approach is especially noteworthy here because I find it so close to our actual experience using GP. Indeed, most colleagues who hear it for the first time utter an inevitable ``didn't we already know this?'' I think the distinction Pickering's approach is able to highlight is exactly the problematic one in our work: between the illusory (but publishable) narrative of ``scientific method'', and the realized experience we have of \emph{performing science}. At the cost of glossing a lot of his well-considered structure, let me summarize.

First, the act of ``doing science'' is \emph{at no point whatsoever} to be understood as one in which an isolated ``researcher'' works in an objective and static field of ``externalities'' and ``facts''. Rather, research begins only when the researcher undertakes to \emph{makes some artifact}: a block of code, a technical instrument, an equation, a sketch, a maquette or even a thoughtful conversation at a conference. Everything before the researcher begins to make these things is a matter of building ``vision'' (my term); research proper only occurs when work is done in the real world. I'll call this work product \emph{the thing made}, and keep in mind that it is a proxy for the vague collection we might otherwise call ``the project''.

We see Pickering's model moving quickly away from more traditional views of science when we realize he has given the inanimate \emph{thing made} an agency of its own. In any real project, we perceive the \emph{thing made} resisting---whether we imagine that it resists ``in itself'', or as an agent of externalities that impinge on the work in progress to make our driving vision differ from reality. The \emph{thing made} comes to represent, in other words, the facts of the actual world, the cultural assumptions and norms of one's discipline, the raw materials and toolkit available to a practitioner, and so forth. ``Resistance'' here is not merely a problem with a software bug or lost raw materials, but includes one's sense \emph{on seeing it} that something's not quite right. The \emph{thing made}'s resistance, as perceived by the researcher, leads her to react in turn.

Consider the language we use in the face of this resistance: it ``feels wrong''; it ``points something out''; it ``wants to do X instead of Y''; it's ``doing something too complicated for me to understand right now''.

And---assuming science, engineering, art or any other creative process is indeed what's being done---the researcher \emph{changes in response to this resistance}. That vision changes, the plan adapts, or in some other way the \emph{thing made} causes a response in the state of the researcher herself. Pickering's Mangle is this emergent dance of inanimate agency: the researcher starting to follow a vision by making (or altering) a \emph{thing}, and the \emph{thing made} in turn acting as a channel for the world itself to steer the researcher in another direction.

Pickering's ``mangle''\footnote{The word ``mangle'' he has chosen is itself interesting and insightful:

\begin{quote}
\ldots{}I find ``mangle'' a convenient and suggestive shorthand for the dialectic because, for me, it conjures up the image of the unpredictable transformations worked upon whatever gets fed into the old-fashioned device of the same name used to squeeze the water out of the washing. It draws attention to the emergently intertwined delineation and reconfiguration of machinic captures and human intentions, practices, and so on. The word ``mangle'' can also be used appropriately in other ways, for instance as a verb. Thus I say that the contours of material and social agency are mangled in practice, meaning emergently transformed and delineated in the dialectic of resistance and accommodation\ldots{}.
\end{quote}} is this emergent dance of agency, undertaken between the researcher and the project: the researcher makes, the \emph{thing made} resists, the researcher is influenced and redirected, \emph{accommodating} the resistance. It may seem glib to say that ``no plan survives contact with the enemy,'' but Pickering's narrative of creative work emphasizes the fact that no \emph{revised} plan survives unchanged, either. In the traditional narrative, we elide the work as it unfolded and re-frame it as a sort of idealized, apersonal Platonic truth: we use the passive voice, we hide the missteps and confusion, we paint a story moving from vision to plan to success.

In a GP setting, the many modes of resistance we perceive are the very ``symptoms'' Rick mused about. Even though we as researchers know we've written all the code and set all the parameters, we're nonetheless willing to speak of a GP run ``doing'' things, as opposed to merely unfolding according to our plan. A GP system does not ``resist'' by, for example, ``having the wrong population size''. Rather it resists by \emph{causing concern or dissatisfaction in the user}, which in turn sparks in that user a practical (if only explanatory) response, which leads them to change the \emph{thing made}.

The broader field of machine learning seems to take a much more traditional stance towards its subject matter: machine learning frameworks (excepting GP) are each discrete tools aimed at producing standardized and reproducible results to particular statistical questions. The result of training a neural network or even a random forest on a given data set is not expected to be a \emph{surprise} in any real sense, but rather the reliable and robust end-product of applying numerical optimization to a well-specified mathematical programming problem. Whether one describes these machine learning processes as ``minimizing out-of-sample error'' or ``maximizing information gain'', the supposed strength of most machine learning approaches is the \emph{unsurprising} nature of their use cases and outputs.

On the other hand, GP has the capacity to \emph{tell us stories}---even in the relatively ``simple'' domain of symbolic regression. The space under consideration by GP is not merely a vector of numerical constants or a binary mask over a suite of input variables, but the \emph{power-set} of inputs, functions over inputs, and higher-order functions over those. We who work in the field can be glib about the ``open-endedness'' of GP systems, but that open-endedness puts them at odds with their supposed relatives in machine learning. While GP \emph{can} be used to explore arbitrarily close to some paradigmatic model, its more typical use case leads to the production of \emph{unexpected} insights---to the degree that a number of us feel justified in treating it as a strong candidate for ``real'' artificial intelligence.

I argue we have that leeway because of the way GP surfaces Pickering's Mangle. When the methodology ``works'', it does so by offering \emph{helpful resistance} in our engagement with the problem at hand, whether in the form of surprising answers, or validation of our suspicions, or simply legible suggestions of ways to make subsequent moves. GP \emph{dances} with us, while most other machine learning methods are ``mere tools''.

Nonetheless, there seems to be a widespread desire, inside and outside our field, to frame GP as a methodology for producing \emph{unsurprising} models from data, more in keeping with the traditional linear of scientific work. That is, an idealized user is expected to proceed something like the users of any other machine learning or mathematical programming framework:

\begin{enumerate}
\item frame your problem in the correct formal language
\item ``get'' a GP system
\item run GP ``on your data''
\item (whatever this is, it's not our problem)
\item you have solved your problem
\end{enumerate}

This is not original with the GP community; there is strong pressure from our peers in other disciplines and our users to promote this use case, not least because it is \emph{exactly} the stance expected in any planning or public policy setting, or in any scientific or programming project management setting. That is, we are under tremendous social pressure to treat GP as a \emph{tool} to be invoked in a known, predictable and well-described planning situation.

The resulting resistance from early-adopting users shouldn't be unexpected. \emph{Being surprising} may well be the worst conceivable behavior for any tool to be used in a traditional project setting. And given that pressure, it's no wonder that so much of GP research is focused on constraining tweaks to bring GP ``into line''. If only GP could be ``tamed'' or made ``adaptive'' so that step (4) above \emph{never happens}.

I imagine this is why so many GP research projects strive for rigor in the form of counting replicates which ``find the solution'': they aim not to convince users of the known strengths of GP, but rather demonstrate to critical peers that GP can be ``tamed'' into a mere tool.

What would a ``replicate'' stand for, to a user who sought to exploit GP's strength in a project (whether theoretical or practical) where \emph{search} rather than the algorithm itself is the primary focus? Projects which authentically ``use'' GP \emph{must necessarily} be those searching for noteworthy answers---which is to say \emph{surprising} and \emph{interesting} answers---that they could not otherwise obtain. Thus, we should better think of a ``replicate'' as a sort of proxy for user frustration in step (4) above: that is, it represents a project in which search begins, stalls, and for which the user cannot see a way to move search forward towards more interesting and useful answers.

But I think we would agree: it is a poor researcher who, when faced with a stumbling block in the form of a black box's misbehavior, doesn't attempt to work around that obstacle. Not to \emph{begin the project again from scratch}, but to make changes and continue. Any researcher who is using GP \emph{realistically} will be watching, and adjusting, and engaging and interacting with the process of search itself.

When a GP user is viewed as working within the traditional \emph{linear} narrative of research, we see her run a population of 100 individuals for 100 generations, peer at the results that have been dumped into a CSV file, and find them wanting. Desiring an actual \emph{answer}, she adjusts the GP parameters and begins another run of 100 generations\ldots{} and repeats as necessary.

But there is \emph{no discernible difference} when we frame this same process of ``many runs'' as a single project involving initial researcher moves, resistances thrown up by the \emph{thing made}, and subsequent accommodations made by the researcher to the new perceived truths. Except, that is, in the researcher's view of her own response to resistance: if she comes to the project with \emph{plans} for running ``ten replicates'', we can only assume she has learned from someone that a search algorithm is \emph{supposed} to forget everything it has learned every 100 generations\ldots{}.

I cannot help but be reminded of the fallacy, surprisingly common both in and outside of our field, that ``artificial intelligence'' must somehow be a self-contained and non-interactive process. That is, that an ``AI candidate'' loses all authenticity as soon as it is ``tweaked'' or ``adjusted'' in the course of operation. It is as if every new-born ``AI'' must be jammed into an air-tight computational container and isolated \emph{until it learns to reason}, and that without exceeding a finite computational budget. If humans creating \emph{real} intelligences treated them anything like the way computer scientists insist we treat nascent \emph{artificial} ones, murder charges would be forthcoming.

There are several practical reasons for us to try something different.

Consider our hypothetical GP user. In keeping with the Behaviorist standards of GP, she is carefully ``not interfering'' with any given run of 100 generations; she can only peer at a results file after the fact. During the course of any 100 generations, all sorts of dynamics have happened: crossover, mutation, selection, all the many random choices. Imagine for a moment we were given perfect access to the entire dynamical pedigree of the unsatisfying results she receives at the end, and were able to backtrack to any point in the run and change a single decision. Before that point, it's unclear how badly things will actually turn out at the 100-generation mark; at some point after that juncture, it's obvious to anybody watching that the whole thing's a mess.

It is, I think, because the myths of artificial intelligence and the linear narrative of science are so deeply intertwined. It is frowned upon to admit in a scientific paper, even when no mistakes were made, that the original vision and plan changed over the course of the project; rather we are expected to describe research \emph{results} as the inevitable outcomes of an ahistorical process, and erase all resistance and accommodation actually done by human beings in the context of their projects. Similarly, it feels somehow wrong to admit in a GP project, even if every parameter was set correctly, that the original vision and plan gave way to the inevitable surprises thrown up by GP's inherent tendencies to do just that.

But insofar as GP \emph{surprises us}, and since that is its sole strength over more predictable and manageable frameworks, we must inevitably see a good fraction of those surprises at least in part as \emph{disappointments} rather than encouraging opportunities to change our plans.  Let's learn new ways to accommodate those disappointments, and stop trying to make them go away.

@@

As far as I can tell, Keith Braithwaite first described his \href{http://cumulative-hypotheses.org/2011/08/30/tdd-as-if-you-meant-it/}{training exercise} for software developers in 2009. The target of the exercise is ``Pseudo-TDD'': the noted habit among software developers who claim to ``know'' and ``do'' test-driven development as part of their daily work towards a sort of thoughtless approximation of the technique.

I should note that a number of agile software development practices share informative relations to genetic programming's dynamics\footnote{I imagine there is an Engineering Studies thesis in this for some aspiring graduate student: Genetic programming and agile development practices arose in the same period and more or less the same culture, and both informed by the same currents in complex systems and emergent approaches to problem-solving.}, but in this work I'll focus on those of TDD. In particular, test-driven development (or more accurately ``test-driven design'') \emph{when done correctly} can break down the complex design space of a software project into a value-ordered set of incremental test cases, focus the developers' attention on those cases alone, inhibit unnecessary ``code bloat'' and feature creep, and produce low-complexity understandable and maintainable software.

TDD \emph{as such} is a rigorous process, to the point where it can be described as ``painful'' (though also ``useful'') by experienced programmers. The steps are deceptively easy to trivialize and misunderstand, especially for those whose habits of thinking about code are ingrained:

\begin{enumerate}
\item Add a little (failing) test which exercises the next behavior you want to build into your codebase
\item Run all tests, expecting only the newest to fail
\item Make the minimal change to your codebase that permits the new test to pass
\item Run all tests, expecting them all to succeed
\item Refactor codebase to remove duplication
\end{enumerate}

Each stage offers a stumbling block for an experienced programmer, but the most salient for us now is the iterative flow of implementation (or ``design'') that it imposes: Each cycle begins with a choice of \emph{which little test should next be added}; each cycle ends with a rigorous process of refactoring, not just of the new code but of the \emph{entire cumulative codebase} produced so far. The middle three steps---implementing a \emph{single} failing test and modifying the codebase \emph{by just enough} so that all tests pass---feel when one is working through them as if they could be automated easily. The \emph{mindfulness} of the process lives in the choice of next steps and (though somewhat less so) of standard refactoring operations.

Braithwaite's exercise does an interesting thing to surface the formal rigor of this approach. In it, the participants (willing, of course, because the exercise is a \emph{kata} or ``refresher'' for experienced software developers to hone their skills) are asked to implement a nominally simple project like the game of Tic-Tac-Toe, given an \emph{ordered} list of features to implement and the artificial restriction that they must go farther than normal TDD practice asks. Rather than producing a suite of tests and a self-contained codebase, they are forced to use \emph{only} refactoring of code added to tests to produce their eventual ``codebase''. In other words, no code can be ``produced'' until the ``smell'' of duplication in the code added to multiple passing tests \emph{provides a warrant} for refactoring it out.

Further, a facilitator patrols ongoing work and deletes \emph{any and all code not called for by a pre-existing failing test}. Words like ``irritating'' and ``annoying'' crop up in participants' accounts of  this onerous backtracking deletion the first few times it happens, as one might imagine. But as Gojko Adzik emphasizes in his descriptions of workshops, the resulting designs for even simple algorithms in this artificially amplified setting seems much more \emph{open-ended} than it would if the software were built under the typical norms and habits an experienced programmer uses in normal conditions.

A number of contextually positive benefits are attributed to agile software development practices, and to TDD within that suite of practices. But the one that brings us here today is that aspect surfaced particularly in Adzik's account of Tic-tac-toe:

@@

In other words: innovative approaches to the problem at hand began to arise, though there wasn't enough time to finish them in the time alloted for the exercise. The analogy to our collective experience to date with genetic programming should start to peek through at this point---though the thoughtful reader will hopefully wonder what utility there is in an analogy drawn between two similarly unsatisfying outcomes.

[more here]

@@

In the same way that Braithwaite's onerous coding exercise is intended to drive the attention of its participants toward test-driven design with its obligation to write ``real'' code \emph{only as a refactoring}, I'd like to be able to demand a \emph{warrant} for every step that moves our changing genetic programming setup away from just plain random guessing. Braithwaite's target of ``Pseudo-TDD'' suggests an analogous ``Pseudo-GP'': one in which the fitness function is the only ``interface'' with the problem itself, and where the representation language, search operators, search objectives and other algorithmic ``parameters'' are \emph{fixed}.\footnote{Braithwaite's participants often acknowledge they \emph{know} and \emph{use} TDD as it's formally described, but rarely take the time to do so unless ``something goes wrong''. I imagine many GP users will say they \emph{know} and \emph{use} all the innumerable design and setup options of GP, but treat them as adjustments to be invoked only when ``something goes wrong''. I offer no particular justification for either anecdote here, but the curious reader is encouraged to poll a sample of participants at any conference (agile or GP).}

Not only do traditional search operators like crossover, mutation and [negative] selection not come ``for free'' in this variant, but in every case we must develop a cogent, data-driven argument in favor of starting them \emph{as part of an ongoing search process}. Similarly, the initial selection criteria will be limited to a minimal subset of the training data, and expansion (and other alterations) of the training set will have to be made in light of measured progress, not assumptions that ``more is better'' in every case.

The result will be an incremental process of refinement of an ongoing search, carried out not at the level of externally-assigned parameter ``tweaks'' but rather by \emph{opening} the black boxes we typically demand and demanding we do surgery to correct their ``pathologies'' (and understand their mysteries) without killing them outright. It is not intended as an ``algorithm'' to supplant those used today, but rather as a forced re-description of what we actually already do.

\subsection{A tableau representation}\hypertarget{a-tableau-representation}{}\label{a-tableau-representation}

The exercise proceeds as a sort of game, in which the User and the System take alternating turns. During the User's turn, she can see state of the system so far and make any of several \emph{moves} from a limited set, each of which involve changing the settings of a \emph{tableau}, which completely describes the state of the search system. During the System's turn, it will execute a finite number of steps in which it creates new individuals. At the end of the System's turn, control reverts to the User, and vice versa.

The tableau, and the suite of moves available to the User, affect three core aspects of search: {\tt operators}, {\tt answers}, and {\tt rubrics}.

@@

An {\tt operator} is any function which takes as argument a (possibly empty) collection of {\tt answers} and produces a new collection of {\tt answers}. Operators thus include ``random guessing'', which in GP systems is often used to build an initial population, any ``crossover'' and ``mutation'' functions. No {\tt operator} can refer to any {\tt answer} not part of its argument set, not can any unset values be set within an {\tt operator}.

So for example, if the User wished an {\tt operator} could be constructed which took 30 {\tt answers} as inputs, read their attributes (including {\tt script} and that subset of their {\tt rubric} values which were already set), and produce a single {\tt answer} in its return set. It could \emph{not} take an {\tt answer}, change it, and keep the new one ``if it was better'', since the new {\tt answer} would have no scores at all.

Note that no process is provided for {\tt answers} to be \emph{removed} from a tableau. The framework is purely cumulative.

All ``parents'' and other {\tt answer} inputs required by an {\tt operator} are chosen by \emph{lexicase selection} automatically, using the complete suite of {\tt rubrics} in play when invoked. Lexicase selection samples each {\tt rubric} with equal probability.

An {\tt answer} might traditionally be called an ``individual'' in GP literature. We can model them programmatically as key--value hash. All new {\tt answers} are born with only a unique {\tt id} and {\tt script} field set. As an {\tt answer} ``matures'' in the unfolding tableau, various other attributes will be set by the System player through the application of {\tt rubrics}.

In the tableau visualization, we represent the unfolding collection of {\tt answers} as the \emph{rows} of a spreadsheet-like table, and their attributes and scores as the \emph{columns}.

@@

A {\tt rubric} is any function which returns a scalar numerical value, given the instantaneous state of the tableau itself. For the most part these values can be understood as ``scores'' for various search objectives, though it is likely that a number of ``implicit objectives'' will also create new {\tt rubrics}.

A {\tt rubric} cannot store intermediate values, but can refer to the values of other {\tt rubric} columns. Thus on specifying a {\tt rubric} like ``sum squared error over a training set'', one must also create  {\tt rubrics} for ``measured error when provided input $i$'' for every element $i$ of the training set. If the training set has 100 elements, then the single SSE {\tt rubric} \emph{implicitly} represents a suite of 101 new columns added to the tableau.

A {\tt rubric} function does however have access to the state of the entire tableau as needed, including ``forcing'' other {\tt rubric} values to be calculated. So it is entirely possible to specify {\tt rubric} functions which score:

>>>>

More than a decade ago, Rick Riolo, Bill Worzel and I were working on a consulting project together that involved genetic programming. As we chatted one day, Rick was asked what he'd most like to see as part of the research program of GP ``moving forward''. Rick's answer informs this contribution, as well as much of my professional work with GP in the years since. As I recall, he said he'd like work to focus on the ``symptoms'' we often see in evolutionary search processes: premature convergence, slow and spotty improvement, the catastrophic lack of diversity, and that ineffable feeling all GP professionals experience when they see results and \emph{know} that somewhere, somebody has Chosen Unwisely.

@@

This contribution aims squarely at an unfortunate sensibility I've detected in our field, which sees the many ``symptoms'' and ``pathologies'' as bugs to be eradicated \emph{before} we let ``real users'' run their own GP systems. Instead I'll argue (belatedly following Rick's lead) that these ``surprises'' and ``disappointments'' are not only inevitable but should be the main focus of our theoretical and practical work. Indeed, much of the value GP can offer can be lost if we approach these phenomena as something to be ``cleaned up'' rather than being \emph{accommodated}.

@@

In hindsight, I've come to think this shocked response was inevitable. It wasn't an unwitting surprise from some silly lay ``customers'' in an application project. Indeed, it's analogous to the way many in our field treat \emph{interesting} GP projects, whether they're ``theoretical'' or ``practical'', ``small'' or ``large''. \emph{Any interesting project will resist our initial plans and expectations.} When we're faced with even a little resistance, our first reaction is to imagine something has ``gone wrong in the setup'', shut the thing off and start again with ``better'' parameters.

It's unusual for anybody to consider interrogating the system \emph{in itself} when it resists, and adapt or accommodate that resistance. Within our field, but also in a surprisingly broad layer of technical society.

@@

Braithwaite's exercise feels subjectively \emph{harder} for more advanced programmers honing their development form; he suggests that novice programmers haven't learned ingrained but questionable habits, and haven't identified ``shortcuts'' that ``simplify'' the practices. In the same way, this exercise will feel \emph{most} artificial and restrictive to those of us with the most experience with GP. But like the martial arts exercises from which software \emph{kata} first were inspired, it isn't intended to be simple or even pleasant for the participants.

I intend it to expose habits I'll tactlessly call ``pseudo-GP'' among those of us who have come to think it's cheap and painless to \emph{just shut it off and start over} when things start to get strange in the course of a GP project. As an important side-effect, it helps surface that philosophical problem I alluded to above. But I have found the most useful result is how strongly and immediately it suggests tools with which one can address Rick's years-ago wish. It forces the participant to formally identify ``pathology'' and ``symptom'' before allowing them to attempt a ``fix''\ldots{} and even that must done with an intentionally limited set of tools.

@@

@@

As far as I can tell, Keith Braithwaite first described his \href{http://cumulative-hypotheses.org/2011/08/30/tdd-as-if-you-meant-it/}{training exercise} for software developers in 2009. His target was a sense that software developers who thought they were using test-driven development practices were in fact doing something more like ``Pseudo-TDD'', a sort of slapdash and habitual approximation lacking many of the benefits of mindful practice.

While I've noted elsewhere that several agile software development practices share useful overlaps with the problems of GP science and engineering\footnote{I imagine there is an Engineering Studies thesis in this for some aspiring graduate student: Genetic programming and agile development practices arose in the same period and more or less the same culture, and both informed by the same currents in complex systems and emergent approaches to problem-solving.}, in this work I'll focus on those of TDD. In particular, the observation that test-driven development (or more accurately ``test-driven design'') \emph{when done correctly} can break down the complex design space of a software project into a value-ordered set of incremental test cases, focus developers' attention on those cases alone, inhibit unnecessary ``code bloat'' and feature creep, and produce low-complexity understandable and maintainable software.

TDD \emph{as such} is a very constraining and rigorous process---to the point where it can easily be described as ``painful'' (though also ``useful'') by experienced programmers. The steps are deceptively easy to gloss, misunderstand or miscommunicate, especially for those whose coding habits are ingrained. To paraphrase Beck and Braithwaite:

\begin{enumerate}
\item Add a little (failing) test which exercises the next behavior you want to build into your codebase
\item Run all tests, expecting \emph{only the newest} to fail
\item Make the minimal change to your codebase that permits the new test to pass
\item Run all tests, expecting them \emph{all} to succeed
\item Refactor the codebase to remove duplication
\end{enumerate}

Even though a single cycle through this iterative process can take less than a minute, each step can throw itself up as a stumbling block for an experienced programmer. But the most salient for us here is the iterative flow of implementation (or ``design'') that the cycle imposes: it begins with a choice of \emph{which little test should next be added}, and ends with a rigorous process of refactoring, not just of the new code but of the \emph{entire cumulative codebase} produced so far. The middle three steps---implementing a \emph{single} failing test and modifying the codebase \emph{by just enough} so that all tests pass---feel when one is working as though they could be automated easily. The \emph{mindfulness} of the process lives in the choice of next steps and (though somewhat less so) of standard refactoring operations.

Braithwaite's exercise does an interesting thing to surface the formal rigor of those decisions, by making them \emph{harder} rather than easier. In ``TDD as if you meant it'', the participants (willing, of course) are asked to implement a nominally simple project like the game of Tic-Tac-Toe using TDD, and are given a list of requisite features and an extra constraint. Rather than using ``normal'' TDD and producing a suite of tests to exercise a separate and self-contained codebase, they are forced to add code \emph{only} to the tests themselves (to make them pass), and can only produce a separate ``codebase'' when duplication or other ``code smells'' drive them to refactor the code already added to tests. In other words, the \emph{demand for a warrant} for writing code is much more stringent.

Throughout the exercise, a facilitator patrols teams of participants and deletes \emph{any and all code not called for by a pre-existing failing test}. Words like ``irritating'' and ``annoying'' crop up in participants' accounts of  this onerous backtracking deletion the first few times it happens, as one might imagine. But as Gojko Adzik emphasizes in descriptions of workshops he's facilitated, the resulting designs even for well-known algorithms in this artificially amplified setting become much more ``open-ended'' than would be expected if the code were written under the offhand attention of an experienced programmer without painful constraints.

Adzik mentions something in his account of a Tic-tac-toe exercise that's especially interesting:

@@

In other words: innovative approaches to the problem at hand began to arise, though there wasn't enough time to finish them in the time alloted for the exercise.

@@

In the same way that Braithwaite's coding exercise uses an onerous extra constraint\footnote{In this the sensibility reminds me of the constraint-driven art collective \href{http://en.wikipedia.org/wiki/Oulipo}{Oulipo}, who are perhaps most famous for the \emph{lipogram}, a literary work which cannot use a particular letter of the alphabet.} to drive participants towards more mindful and insightful decision-making, in this exercise I will demand a \emph{warrant} for each implementation decision that moves a running GP setup away from random guessing. Braithwaite's target of ``Pseudo-TDD'' suggests an analogous ``Pseudo-GP'': one in which the fitness function and \emph{post hoc} analysis is the only ``interface'' with the problem itself, and where the representation language, search operators, search objectives and other algorithmic ``parameters'' are \emph{fixed} in the course of the run.\footnote{Braithwaite's participants often acknowledge they \emph{know} and \emph{use} TDD as it's formally described, but rarely take the time to do so unless ``something goes wrong''. I imagine many GP users will say they \emph{know} and \emph{use} all the innumerable design and setup options of GP, but treat them as adjustments to be invoked only when ``something goes wrong''. I offer no particular justification for either anecdote here, but the curious reader is encouraged to poll a sample of participants at any conference (agile or GP).}

Not only will traditional search operators like crossover, mutation and [negative] selection not come ``for free'' in this variant, but in every case we must develop a cogent argument in favor of starting them \emph{as part of an ongoing search process}. Similarly, the initial selection criteria will be limited to a minimal single element of the training data, and expansion of the training set in use will have to be made in light of measured progress, not the assumption that ``more is better''.

The result is a painfully incremental process of refinement of an unstoppable search which has \emph{intentionally} ``started wrong'', which must be carried out not by stopping the whole thing and launching it again with parameter ``tweaks'' but rather by \emph{doing surgery on the living patient} to correct any perceived ``pathologies''. Along the way, a fraction of the mysteries of ``pathology'' become much clearer and well-defined, even when they cannot be solved easily.

Again, let me emphasize that this is not intended as an ``algorithm'' in its own right, but rather as a suite of constraints imposed on what we do without paying attention already.

\subsection{The game tableau}\hypertarget{the-game-tableau}{}\label{the-game-tableau}

The exercise is structured as a sort of game with two players, one of which is a running GP System, and one which is a human User (or team of Users collaborating on making a single move at a time). The two players make alternating turns.

The game ``setup'' can be imagined as a \emph{tableau} with two components: A single list of search {\tt operators}, and a large (and growing) two-dimensional spreadsheet-like table with {\tt answers} as its row labels and {\tt rubrics} as its column labels.

During the User's turn, she can examine the state of the tableau, including all of its history, and can do any amount of data analysis or statistical consideration she wants. She is however permitted no more than \emph{two} moves: in each turn she can code and launch one new {\tt operator}, and can code and launch one new {\tt rubric}.

Finally, the User is obliged to provide a convincing \emph{warrant} for every move she make, to be reviewed by an external Facilitator. In the event this warrant is seen by the Facilitator to justify the particular move, he has the right to delete all {\tt operators} and {\tt rubrics} added by the User from that point forward \emph{without resetting the System's moves}.

Facilitator buy-in for every move is therefore \emph{strongly} encouraged.

During the System's turn, it will execute a finite number of steps in which it creates new individuals using the specified search {\tt operators} and {\tt rubrics}. At the end of the System's turn, control reverts to the User, and vice versa.

An {\tt answer} might traditionally be called an ``individual'' in the GP literature. It is modeled here as key--value hash (or a simple prototypal object). Every new {\tt answer} is born with only a unique {\tt id} and its {\tt script} field set. As they ``mature'' during the System's turns, various other attributes (``scores'') will be set by {\tt rubrics}.

Answers never ``die'', and cannot be removed from the tableau by either player as search progresses. The User can only reduce or increase the chance that certain types of {\tt answer} contribute to later generations by constructing and launching {\tt rubrics} to adjust selective forces.

@@

An {\tt operator} is a coded function which takes as its argument an unordered  collection of {\tt answers}, and which produces a new collection of {\tt answers} as output.

The initial tableau includes only a single {\tt operator}, which implements a pre-coded ``random guessing'' algorithm. On the User's move, she may build and launch other more complex (and familiar) {\tt operators} like ``crossover'' or ``mutation'' functions. A wide variety of {\tt operator} structures are possible in the domain-specific language provided.

The specific set of {\tt answers} to which an {\tt operator} is applied cannot be chosen directly by the User. All ``parents'' defined for each {\tt operator} are chosen by \emph{lexicase selection} automatically during the System's turns, using the complete suite of {\tt rubrics} in play when invoked. This selection process samples each {\tt rubric} in the tableau with equal probability.

It is feasible for the User to design a new ``selection'' {\tt operator}, but it must rely on the immediate state of the {\tt answers} it is given; in other words, no {\tt operator} can explicitly invoke a particular {\tt rubric} (or vice versa).

Note again that no mechanism exists which removes {\tt answers} from the tableau.

@@

A {\tt rubric} is a function which returns a scalar value (not necessarily a \emph{number}) for a given {\tt answer}, conditioned as needed on the instantaneous tableau state. When a {\tt rubric} is applied to an {\tt answer}, it sets a new value (or ``score'') in that {\tt answer}.

No ``score'' is ever updated in an {\tt anwer} once set by a {\tt rubric}. However, if multiple {\tt answers} exist which have the same {\tt script}, each may be scored at different times or in different contexts, and the resulting values may differ.

An aggregate or \emph{higher-order} {\tt rubric} can be created, but as a consequence it ``entails'' all of its component {\tt rubrics}. So if the User constructs a higher-order {\tt rubric} like ``max error over 30 training cases'', when she launches that {\tt rubric} 31 in total will be launched, with the first 30 being in the form ``absolute error for training case $i$''.

If a User launches such a higher-order {\tt rubric} on her turn, all its entailed {\tt rubrics} will also be added before the System takes its turn.

When one of these ``input'' {\tt rubrics} is applied during the System's turns, it does \emph{not} automatically cause the higher-order aggregate to be applied. However, when the higher-order {\tt rubric} is applied, it will force all of its ``input'' {\tt rubrics} to be calculated and applied first.
